Concurrent Programming with CXL

### 2020-06-23 book/paper.tex
Better exercise numbering

\documentclass{report}
\usepackage{fullpage}
\usepackage{imakeidx}\makeindex
\usepackage{hyperref}
\usepackage[nottoc,notlot,notlof]{tocbibind}
\usepackage[toc]{glossaries}\makeglossaries
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}
\usepackage{fancyvrb}
\usepackage{comment}
\usepackage{amsmath}

@@

\section*{Exercises}
\begin{enumerate}
\item See what happens if, instead of initializing \texttt{result} to 0,
you initialize it to 1.  (You do not need to understand the error report at this time.)
Explain in English what the problem is.
\item Write a CXL program that computes squares by repeated adding.  So the program
should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
\end{enumerate}

@@

\begin{enumerate}
\item CXL programs can usually be easily translated into Python.  For example,
Figure~\ref{fig:incpy} is a Python version of Figure~\ref{fig:inc}.
\begin{enumerate}
\item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run Figure~\ref{fig:incpy} 1000 times.  How many times does the assertion
fail (if any).
\end{enumerate}
\item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
incrementer thread increment \texttt{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
10 times (using Python).
Report how many times the assertion fails and what the value of \texttt{count}
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
\item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes
and run them through CXL.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
\end{enumerate}

@@

\begin{enumerate}
\item Figure~\ref{fig:incenter} shows an attempt at trying to fix the code of
    Figure~\ref{fig:inc}.  Run it through CXL and see what happens.  Based on
    the error output, describe in English what is wrong with the code by describing,
    in broad steps, how running the program can get into a bad state.
\item What if we moved line~2 of Figure~\ref{fig:incenter} to after the \texttt{if}
    statement (between lines~7 and~8)?  Do you think that would work?  Run it through
    CXL and describe either why it works or why it does not work.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now plug the following code \texttt{while True: pass;;}
just before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.
Mutual exclusion is guaranteed but progress is violated.
CXL should print a short trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Figure~\ref{fig:csonebit} presents another correct solution to the
mutual exclusion problem.  It is similar to the one in
Figure~\ref{fig:upflags}, but has a process \emph{back out and try again}
if it finds that the other process is either trying to enter the critical
section or already has.  Compare this algorithm with Peterson's.  What are
the advantages and disadvantages?
\item
Can you find an inductive invariant for the algorithm in
Figure~\ref{fig:csonebit} to prove it correct?
Here's a pseudo-code version of the algorithm to help you.  Each line
is an atomic action:
\begin{code}
\begin{verbatim}
    initially: flagX = flagY = False

    process X:                          process Y:
        X0: flagX = True                   Y0: flagY = True
        X1: if not flagY goto X4           Y1: if not flagX goto Y4
        X2: flagX = False                  Y2: flagY = False
        X3: goto X0                        Y3: goto Y0
        X4: ...critical section...         Y4: ...critical section...
        X5: flagX = False                  Y5: flagY = False
\end{verbatim}
\end{code}
\item A colleague of the author asked if the first two assignments in
Peterson's algorithm (setting \texttt{flags[self]})
to \texttt{True} and \texttt{turn} to \texttt{1 - self}) can be reversed.
After all, they are different variables assigned independent values---in a
sequential program one could surely swap the two assignments.
See if you can figure out for yourself if the two assignments can be
reversed.  Then run the program in Figure~\ref{fig:peterson} after reversing
the two assignments and describe in English what happens.
\item Bonus question:
Can you generalize Peterson's algorithm to more than two processes?
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Implement an atomic swap operation.  It should take two pointer arguments
and swap the values.
\item Implement a spinlock using the atomic swap operation.
\item Write out the invariants that need to hold and check them using CXL.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Write a synchronization module using your own implementation of \texttt{Lock}
(for example, using Peterson's algorithm or using \texttt{swap}).
\item
Run Figure~\ref{fig:incfixed} using (1) \texttt{synch}, (2), \texttt{synchS},
and (3) your own module.  Report how many states were explored by CXL for each
version of the module.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Draw additional states and steps in Figure~\ref{fig:rdwr}
for three processes.
\item Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} processes to acquire it at the same time.  A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
\item Write a test program for your bound lock
that checks that no more than \texttt{M} processes can acquire the
bound lock.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
per process rather than a simple counter?
Can you write a version of this test that uses a counter instead
of a flag per process?
% \item Write a ``library implementation'' of reader/writer locks so multiple
% reader/writer locks can be instantiated.
\item Write a test program for ``bound locks'' (as implemented in the
exercises of the last chapter) that checks that up to \texttt{M} processes
can acquire the bound lock at the same time.
\item It can be useful for a process to \emph{downgrade} its write lock
to a read lock when its done writing but not done reading.  Implement
\texttt{downgrade}.  Note, you cannot do this simply by releasing the
write lock and then obtaining the read lock.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Does it matter in what order the philosophers lay down their forks?
\item Figure~\ref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
processes.  Can you fix the problem?
\end{enumerate}

@@

\section*{Exercices}
\begin{enumerate}
\item Add an assertion that demonstrates that at most three
Italian philosophers are eating at the same time?
\item Implement a test program that checks to see that
the solution allows more than one Italian philosopher to eat at the same time?
\item Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
The array to be sorted is stored in  shared variable \texttt{a}.
Another shared variable, \texttt{todo}, contains the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker processes''
that should replace the \texttt{qsort} code.  Each worker loops until \texttt{todo}
is empty and sorts the ranges that it finds until then.  The \texttt{main}
process needs to wait until all workers are done.  Use a semaphore for this:
the \texttt{main} process should use \texttt{P} to wait for each
worker to terminate.  After that it should check that the array is sorted and
that it is a permutation of the original array.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Change the code in Figure~\ref{fig:boundedbuffer} so that
\texttt{produce()} fails if the buffer is full rather than block.
\texttt{produce()} should return \texttt{False} if it fails
and return \texttt{True} if it succeeds.
\item Implement a \emph{bounded stack}.  Method \texttt{push(item)}
should block if the stack is full.  Method \texttt{pop()} should block
until the stack is non-empty and then return an item.
\item Implement a thread-safe GPU allocator by modifying Figure~\ref{fig:gpu}.
There are \texttt{N} GPUs identified by the numbers
1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
of an available GPU, blocking if there is currently no available GPU.
Method \texttt{gpuRelease(gpu)} releases the given GPU.  It never needs
to block.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Assume that you have only binary semaphores.  Implement counting
semaphores using binary semaphores using split binary semaphores.
\item Implement a solution to the producer/consumer problem
(Figure~\ref{fig:boundedbuffer}) using split binary semaphores.
\item Cornell's campus features some one-lane bridges.  Cars can
only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.  Write methods
\texttt{nb\_enter()} that a car must invoke before going northbound on
the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
for southbound cars.

\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Write a fair solution to the one-lane bridge problem of Chapter~\ref{ch:sbs}.
If you want to use the queue method, you can change the signature of the
methods to enter the bridge, adding an argument that contains a pointer to a
semaphore.
\end{enumerate}

@@

\section*{Exercises}
\begin{enumerate}
\item Write a new version of the GPU allocator of Chapter~\ref{ch:bb}.  In this version,
a process is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A process does not have to return all the GPUs it allocated at once.
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\end{enumerate}

>>>>

\documentclass{report}
\usepackage{fullpage}
\usepackage{imakeidx}\makeindex
\usepackage{hyperref}
\usepackage[nottoc,notlot,notlof]{tocbibind}
\usepackage[toc]{glossaries}\makeglossaries
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}
\usepackage{fancyvrb}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{etoolbox}

\AtBeginEnvironment{enumerate}{\everymath{\displaystyle}}
\newlist{problems}{enumerate}{2}
\setlist[problems]{wide=0pt}
\setlist[problems,1]{label=\thechapter.\arabic*, font=\bfseries, wide=0pt}
\setlist[problems,2]{label=(\alph*), wide =0.5em, topsep=2pt, itemsep=2pt}

@@

\section*{Exercises}
\begin{problems}
\item See what happens if, instead of initializing \texttt{result} to 0,
you initialize it to 1.  (You do not need to understand the error report at this time.)
Explain in English what the problem is.
\item Write a CXL program that computes squares by repeated adding.  So the program
should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
\end{problems}

@@

\begin{problems}
\item CXL programs can usually be easily translated into Python.  For example,
Figure~\ref{fig:incpy} is a Python version of Figure~\ref{fig:inc}.
\begin{enumerate}
\item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run Figure~\ref{fig:incpy} 1000 times.  How many times does the assertion
fail (if any).
\end{enumerate}
\item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
incrementer thread increment \texttt{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
10 times (using Python).
Report how many times the assertion fails and what the value of \texttt{count}
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
\item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes
and run them through CXL.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
\end{problems}

@@

\begin{problems}
\item Figure~\ref{fig:incenter} shows an attempt at trying to fix the code of
    Figure~\ref{fig:inc}.  Run it through CXL and see what happens.  Based on
    the error output, describe in English what is wrong with the code by describing,
    in broad steps, how running the program can get into a bad state.
\item What if we moved line~2 of Figure~\ref{fig:incenter} to after the \texttt{if}
    statement (between lines~7 and~8)?  Do you think that would work?  Run it through
    CXL and describe either why it works or why it does not work.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now plug the following code \texttt{while True: pass;;}
just before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.
Mutual exclusion is guaranteed but progress is violated.
CXL should print a short trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Figure~\ref{fig:csonebit} presents another correct solution to the
mutual exclusion problem.  It is similar to the one in
Figure~\ref{fig:upflags}, but has a process \emph{back out and try again}
if it finds that the other process is either trying to enter the critical
section or already has.  Compare this algorithm with Peterson's.  What are
the advantages and disadvantages?
\item
Can you find an inductive invariant for the algorithm in
Figure~\ref{fig:csonebit} to prove it correct?
Here's a pseudo-code version of the algorithm to help you.  Each line
is an atomic action:
\begin{code}
\begin{verbatim}
    initially: flagX = flagY = False

    process X:                          process Y:
        X0: flagX = True                   Y0: flagY = True
        X1: if not flagY goto X4           Y1: if not flagX goto Y4
        X2: flagX = False                  Y2: flagY = False
        X3: goto X0                        Y3: goto Y0
        X4: ...critical section...         Y4: ...critical section...
        X5: flagX = False                  Y5: flagY = False
\end{verbatim}
\end{code}
\item A colleague of the author asked if the first two assignments in
Peterson's algorithm (setting \texttt{flags[self]})
to \texttt{True} and \texttt{turn} to \texttt{1 - self}) can be reversed.
After all, they are different variables assigned independent values---in a
sequential program one could surely swap the two assignments.
See if you can figure out for yourself if the two assignments can be
reversed.  Then run the program in Figure~\ref{fig:peterson} after reversing
the two assignments and describe in English what happens.
\item Bonus question:
Can you generalize Peterson's algorithm to more than two processes?
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Implement an atomic swap operation.  It should take two pointer arguments
and swap the values.
\item Implement a spinlock using the atomic swap operation.
\item Write out the invariants that need to hold and check them using CXL.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Write a synchronization module using your own implementation of \texttt{Lock}
(for example, using Peterson's algorithm or using \texttt{swap}).
\item
Run Figure~\ref{fig:incfixed} using (1) \texttt{synch}, (2), \texttt{synchS},
and (3) your own module.  Report how many states were explored by CXL for each
version of the module.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Draw additional states and steps in Figure~\ref{fig:rdwr}
for three processes.
\item Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} processes to acquire it at the same time.  A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
\item Write a test program for your bound lock
that checks that no more than \texttt{M} processes can acquire the
bound lock.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
per process rather than a simple counter?
Can you write a version of this test that uses a counter instead
of a flag per process?
% \item Write a ``library implementation'' of reader/writer locks so multiple
% reader/writer locks can be instantiated.
\item Write a test program for ``bound locks'' (as implemented in the
exercises of the last chapter) that checks that up to \texttt{M} processes
can acquire the bound lock at the same time.
\item It can be useful for a process to \emph{downgrade} its write lock
to a read lock when its done writing but not done reading.  Implement
\texttt{downgrade}.  Note, you cannot do this simply by releasing the
write lock and then obtaining the read lock.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Does it matter in what order the philosophers lay down their forks?
\item Figure~\ref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
processes.  Can you fix the problem?
\end{problems}

@@

\section*{Exercices}
\begin{problems}
\item Add an assertion that demonstrates that at most three
Italian philosophers are eating at the same time?
\item Implement a test program that checks to see that
the solution allows more than one Italian philosopher to eat at the same time?
\item Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
The array to be sorted is stored in  shared variable \texttt{a}.
Another shared variable, \texttt{todo}, contains the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker processes''
that should replace the \texttt{qsort} code.  Each worker loops until \texttt{todo}
is empty and sorts the ranges that it finds until then.  The \texttt{main}
process needs to wait until all workers are done.  Use a semaphore for this:
the \texttt{main} process should use \texttt{P} to wait for each
worker to terminate.  After that it should check that the array is sorted and
that it is a permutation of the original array.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Change the code in Figure~\ref{fig:boundedbuffer} so that
\texttt{produce()} fails if the buffer is full rather than block.
\texttt{produce()} should return \texttt{False} if it fails
and return \texttt{True} if it succeeds.
\item Implement a \emph{bounded stack}.  Method \texttt{push(item)}
should block if the stack is full.  Method \texttt{pop()} should block
until the stack is non-empty and then return an item.
\item \label{ex:gpu} Implement a thread-safe GPU allocator by modifying
Figure~\ref{fig:gpu}.
There are \texttt{N} GPUs identified by the numbers
1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
of an available GPU, blocking if there is currently no available GPU.
Method \texttt{gpuRelease(gpu)} releases the given GPU.  It never needs
to block.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Assume that you have only binary semaphores.
Implement counting semaphores using binary semaphores using split binary semaphores.
\item Implement a solution to the producer/consumer problem
(Figure~\ref{fig:boundedbuffer}) using split binary semaphores.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
Cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.  Write methods
\texttt{nb\_enter()} that a car must invoke before going northbound on
the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
for southbound cars.

\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Write a fair solution to the one-lane bridge problem of
Exercise~\ref{ex:onelane}.
If you want to use the queue method, you can change the signature of the
methods to enter the bridge, adding an argument that contains a pointer to a
semaphore.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Write a new version of the GPU allocator in Exercise~\ref{ex:gpu}.
In this version,
a process is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A process does not have to return all the GPUs it allocated at once.
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\end{problems}
